---
title: "FDRI GEAR-Hourly Zarr in R"
subtitle: "Using the native `zarr` R package to explore gridded rainfall data"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 9
    fig-height: 5
    self-contained: true
    link-external-icon: true
    link-external-newwindow: true
execute:
  echo: true
  warning: false
  message: false
  cache: false
engine: knitr
---

::: {.callout-note}
## Launch this notebook in other environments

| | |
|---|---|
| [![Open In Colab](https://colab.research.google.com/assets/colab-badge.svg)](https://colab.research.google.com/github/YOUR_ORG/YOUR_REPO/blob/main/fdri_gearhrly_zarr.ipynb) | Google Colab |
| [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/YOUR_ORG/YOUR_REPO/HEAD?labpath=fdri_gearhrly_zarr.ipynb) | Binder |
| [![NERC DataLabs](https://img.shields.io/badge/Open%20in-NERC%20DataLabs-blue?logo=jupyter)](https://datalabs.nerc.ac.uk) | NERC DataLabs |

Replace `YOUR_ORG/YOUR_REPO` with your GitHub path.
:::

## About

This notebook explores **GEAR-Hourly** (Gridded Estimates of Areal Rainfall — Hourly), a
gridded rainfall dataset for Great Britain produced as part of the
[FDRI (Floods and Droughts Research Infrastructure)](https://fdri.org.uk) project,
hosted on JASMIN S3 object storage.

We use the **[`zarr`](https://r-cf.github.io/zarr/)** package — a native, extensible R
driver for Zarr v3 — to open and explore the store directly over HTTP.

```
https://fdri-o.s3-ext.jc.rl.ac.uk/gearhrly/gearhrly_15day_100km_chunks.zarr
```

---

## 0. Setup

```{r setup}
pkgs <- c("zarr", "curl", "ggplot2", "dplyr", "tidyr", "sf",
          "rnaturalearth", "rnaturalearthdata", "scales", "lubridate")
new  <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new) > 0) install.packages(new, repos = "https://cloud.r-project.org")

suppressPackageStartupMessages({
  library(zarr)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(sf)
  library(rnaturalearth)
  library(rnaturalearthdata)
  library(scales)
  library(lubridate)
})

cat("zarr version:", as.character(packageVersion("zarr")), "\n")
```

---

## 1. Open the Zarr store

```{r open-store}
store_url <- "https://fdri-o.s3-ext.jc.rl.ac.uk/gearhrly/gearhrly_15day_100km_chunks.zarr"

z <- open_zarr(store_url)
print(z)
```

---

## 2. Inspect structure and metadata

### Store hierarchy

```{r hierarchy}
z$hierarchy()
```

### List arrays

```{r list-arrays}
cat("Arrays:\n")
cat(paste(" •", z$arrays), sep = "\n")
cat("\nGroups:\n")
cat(paste(" •", z$groups), sep = "\n")
```

### Array details

```{r array-details}
for (arr_path in z$arrays) {
  arr <- z[[arr_path]]
  cat(sprintf("\n── %s ──\n", arr_path))
  cat(sprintf("  Shape  : %s\n", paste(arr$shape, collapse = " × ")))
  cat(sprintf("  Chunks : %s\n", paste(arr$chunk_grid$chunk_shape, collapse = " × ")))
  cat(sprintf("  dtype  : %s\n", arr$data_type$data_type))
  attrs <- arr$attributes
  if (length(attrs) > 0) {
    for (nm in names(attrs)) {
      val <- paste(unlist(attrs[[nm]]), collapse = ", ")
      if (nchar(val) > 80) val <- paste0(substr(val, 1, 77), "...")
      cat(sprintf("  %-20s : %s\n", nm, val))
    }
  }
}
```

---

## 3. Read coordinates

```{r read-coords}
read_coord <- function(z, path) {
  tryCatch(as.numeric(z[[path]][]), error = function(e) {
    message("Could not read '", path, "': ", conditionMessage(e)); NULL
  })
}

lons  <- read_coord(z, "/longitude")
lats  <- read_coord(z, "/latitude")
times <- read_coord(z, "/time")

# Decode time
time_arr   <- z[["/time"]]
time_attrs <- time_arr$attributes
time_units <- time_attrs[["units"]]
cat("Time units:", time_units, "\n")

origin_str <- sub(".*since\\s+", "", time_units)
origin_dt  <- as.POSIXct(origin_str, tz = "UTC",
  format = ifelse(grepl(":", origin_str), "%Y-%m-%d %H:%M:%S", "%Y-%m-%d"))
unit_word  <- tolower(sub("\\s+since.*", "", time_units))
multiplier <- switch(unit_word, hours = 3600, days = 86400, minutes = 60, seconds = 1, 3600)
datetimes  <- origin_dt + as.numeric(times) * multiplier

cat(sprintf("Lons  : %d values, %.3f° to %.3f°\n", length(lons), min(lons), max(lons)))
cat(sprintf("Lats  : %d values, %.3f° to %.3f°\n", length(lats), min(lats), max(lats)))
cat(sprintf("Times : %d steps, %s to %s\n",
  length(datetimes), format(min(datetimes)), format(max(datetimes))))
```

---

## 4. Set data variable

```{r choose-var}
data_var   <- "rainfall_amount"   # adjust if needed
data_arr   <- z[[paste0("/", data_var)]]
data_attrs <- data_arr$attributes
var_units  <- data_attrs[["units"]] %||% ""
var_long   <- data_attrs[["long_name"]] %||% data_attrs[["standard_name"]] %||% data_var
cat("Variable:", data_var, "|", var_long, "|", var_units, "\n")
```

---

## 5. Time series at a point

```{r read-ts}
target_lon <- -3.19
target_lat <- 55.95

i_lon <- which.min(abs(lons - target_lon))
i_lat <- which.min(abs(lats - target_lat))

cat(sprintf("Nearest cell: lon=%.3f (idx %d), lat=%.3f (idx %d)\n",
  lons[i_lon], i_lon, lats[i_lat], i_lat))

n_time  <- length(times)
ts_raw  <- data_arr[1:n_time, i_lat, i_lon]
ts_df   <- data.frame(datetime = datetimes, value = as.numeric(ts_raw))
```

### Hourly time series

```{r plot-ts, fig.height=4}
ggplot(ts_df, aes(x = datetime, y = value)) +
  geom_line(colour = "#1d6fa4", linewidth = 0.5, alpha = 0.8) +
  scale_x_datetime(labels = date_format("%b %Y")) +
  labs(
    title    = paste("FDRI GEAR-Hourly:", var_long),
    subtitle = sprintf("lon=%.3f°, lat=%.3f°", lons[i_lon], lats[i_lat]),
    x = NULL, y = paste0(var_long, " (", var_units, ")"),
    caption  = "Source: FDRI / UKCEH"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

### Daily totals

```{r plot-daily, fig.height=4}
ts_df |>
  mutate(date = as.Date(datetime)) |>
  group_by(date) |>
  summarise(total = sum(value, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = date, y = total)) +
  geom_col(fill = "#1d6fa4", alpha = 0.8) +
  scale_x_date(labels = date_format("%b %Y")) +
  labs(
    title    = paste("Daily total:", var_long),
    subtitle = sprintf("lon=%.3f°, lat=%.3f°", lons[i_lon], lats[i_lat]),
    x = NULL, y = paste0("Daily total (", var_units, ")"),
    caption  = "Source: FDRI / UKCEH"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

---

## 6. Spatial map

```{r spatial-map, fig.height=7}
t_index    <- 1
slice_time <- format(datetimes[t_index], "%Y-%m-%d %H:%M UTC")

slice_raw <- data_arr[t_index, 1:length(lats), 1:length(lons)]
slice_mat <- matrix(as.numeric(slice_raw), nrow = length(lats), ncol = length(lons))

map_df        <- expand.grid(lon = lons, lat = lats)
map_df$value  <- as.vector(t(slice_mat))

gb <- ne_countries(country = c("United Kingdom", "Ireland"),
                   scale = "medium", returnclass = "sf")

ggplot() +
  geom_raster(data = map_df, aes(x = lon, y = lat, fill = value), na.rm = TRUE) +
  geom_sf(data = gb, fill = NA, colour = "grey20", linewidth = 0.3) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1,
    name = var_units, na.value = "grey90", trans = "sqrt", labels = comma) +
  coord_sf(xlim = range(lons), ylim = range(lats), expand = FALSE) +
  labs(
    title    = paste("FDRI GEAR-Hourly:", var_long),
    subtitle = slice_time,
    x = "Longitude", y = "Latitude",
    caption  = "Source: FDRI / UKCEH"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right",
        panel.grid = element_line(colour = "grey80", linewidth = 0.2))
```

---

## 7. Accumulated rainfall map

```{r accum-map, fig.height=7}
n_steps     <- min(24, n_time)
window_raw  <- data_arr[1:n_steps, 1:length(lats), 1:length(lons)]
window_arr  <- array(as.numeric(window_raw), dim = c(n_steps, length(lats), length(lons)))
total_grid  <- apply(window_arr, c(2, 3), sum, na.rm = TRUE)

total_df       <- expand.grid(lon = lons, lat = lats)
total_df$value <- as.vector(t(total_grid))

ggplot() +
  geom_raster(data = total_df, aes(x = lon, y = lat, fill = value), na.rm = TRUE) +
  geom_sf(data = gb, fill = NA, colour = "grey20", linewidth = 0.3) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1,
    name = paste0("Total\n(", var_units, ")"), na.value = "grey90",
    trans = "sqrt", labels = comma) +
  coord_sf(xlim = range(lons), ylim = range(lats), expand = FALSE) +
  labs(
    title    = paste0("FDRI GEAR-Hourly: ", n_steps, "-hour accumulated ", var_long),
    subtitle = sprintf("%s to %s",
      format(datetimes[1], "%Y-%m-%d %H:%M UTC"),
      format(datetimes[n_steps], "%Y-%m-%d %H:%M UTC")),
    x = "Longitude", y = "Latitude",
    caption  = "Source: FDRI / UKCEH"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right",
        panel.grid = element_line(colour = "grey80", linewidth = 0.2))
```

---

## Session info

```{r session}
sessionInfo()
```
